Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-27 mr-review.Rmd
added text about MR GENIUS

Complex trait genetics studies in humans have yielded thousands of new associations for  
diverse traits. The next step in understanding biology of health and disease is to
identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
in genome-wide association studies can uncover causal relationships 
between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
the exposure complex trait, investigators can tease out causal effects from observational data, provided
that three assumptions are satisfied. We discuss below the growth of genome-wide association studies 
in parallel with development of Mendelian randomization methods. We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.

@@

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Two-sample 
Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian]. They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

@@

Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
publication of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to two challenges:

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping.

@@

InSIDE assumption

@@

Limited information maximum likelihood is sometimes called the maximum likelihood 
counterpart of two-stage least squares, and it yields the same causal estimate as 
two-stage least squares and the ratio method when used with a single SNP 
instrument. Additionally, the limited information maximum likelihood framework can accommodate $K$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.

@@

For each subject $i$, the exposure and outcome values come from a bivariate normal 
distribution. They assumed that the mean of the exposure distribution is linear in 
the SNP instrument and that the mean of the outcome distribution is a linear in the 
mean exposure [@burgess_review_2017;@jones_choice_2012]. 
@burgess_improving_2012 demonstrated that the Bayesian model, with weak 
instruments, outperforms the limited information maximum likelihood model in terms 
of frequentist coverage levels.

@@

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in MR studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

@@

In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity Q test. In a meta-analysis of MR Wald estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).

@@

@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
weighted median estimator

@@

@windmeijer_use_2019, building on the research from 
@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
after recognizing that sisVIVE misclassifies valid SNP instruments as 
invalid when the invalid SNP instruments have strong effects on the 
exposure. Consistent selection of invalid SNP instruments, they found, 
depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
proposed a median estimator with consistency that doesn't 
depend on SNP instrument-exposure association strength or the SNP instrument 
correlation structure. They then applied methods from @zou2006adaptive to 
achieve a consistent estimator with the same asymptotic distribution as the oracle 
two-stages least squares.

@@

@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
developing three new methods for two-sample MR analysis with GWAS summary data: robust 
regression, penalized weights, and LASSO penalization. The first two can be viewed as 
modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
downweighting or excluding variants with hetero-
geneous causal estimates.
Recall that MR-Egger offers consistent causal effect estimates when there is no 
horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
formulate an estimator with high breakdown point and high efficiency. The second estimation 
method, with penalized weights, is much like the penalized median estimator, from above, 
except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
objective function from MR-Egger to get Equation \@ref(eq:lasso) as the needed objective function.

@@

@bowden_improving_2019

@@

@foley_mr-clust_2021

@tchetgen_tchetgen_genius_2021 MR-GENIUS

>>>>

Complex trait genetics studies in humans have yielded thousands of new associations for  
diverse traits. The next step in understanding biology of health and disease is to
identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
in genome-wide association studies can uncover causal relationships 
between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
the exposure complex trait, investigators can tease out causal effects from observational data, provided
that necessary assumptions are satisfied. We discuss below the growth of genome-wide association studies 
in parallel with development of Mendelian randomization methods. We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.

@@

Great interest has developed in inferring causal relationships between complex 
traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
enhancing our understanding of the biology of health and disease. Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].

They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

@@

Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
publication of the early GWAS (Figure \@ref(fig:article-counts)). Many studies use the "two-sample" MR design, in which one cohort of subjects has measurements for the exposure, while a second cohort has measurements for a second trait [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

Inverse variance 
weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to two challenges:

@@

Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
be statistically independent. This restriction to 
independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
selected through linkage disequilibrium (LD) clumping [@purcell2007plink].

@@

@@

Limited information maximum likelihood is sometimes called the maximum likelihood 
counterpart of two-stage least squares, and it yields the same causal estimate as 
two-stage least squares and the ratio method when used with a single SNP 
instrument. Additionally, the limited information maximum likelihood framework can accommodate $K > 1$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.

@@

For each subject $i$, the exposure and outcome values come from a bivariate normal 
distribution. The mean of the exposure distribution is assumed linear in 
the SNP instrument, and the mean of the outcome distribution is linear in the 
mean exposure [@burgess_review_2017;@jones_choice_2012]. 
@burgess_improving_2012 demonstrated that the Bayesian model, with weak 
instruments, outperforms the limited information maximum likelihood model in terms 
of frequentist coverage levels.

@@

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
in these studies. The solution to this task differs among published methods. Early methods chose a 
set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
this is the finding that including multiple correlated SNPs in transcriptome-wide association
studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].

@@

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

@@

In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity Q test. In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).

@@

@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald 
ratio estimates.

@@

@windmeijer_use_2019, building on the research from 
@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
after recognizing that sisVIVE misclassifies valid SNP instruments as 
invalid when the invalid SNP instruments have strong effects on the 
exposure. Consistent selection of invalid SNP instruments, they found, 
depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
proposed a median estimator with consistency that doesn't 
depend on SNP instrument-exposure association strength or the SNP instrument 
correlation structure. They then applied methods from @zou2006adaptive to 
achieve a consistent estimator with the same asymptotic distribution as the oracle 
two-stages least squares. One important limitation of this thread of research is that
both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.

@@

@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
developing three new methods for two-sample MR analysis with GWAS summary data: robust 
regression, penalized weights, and LASSO penalization. The first two can be viewed as 
modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
downweighting or excluding variants with heterogeneous causal estimates.
Recall that MR-Egger offers consistent causal effect estimates when there is no 
horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
formulate an estimator with high breakdown point and high efficiency. The second estimation 
method, with penalized weights, is much like the penalized median estimator, from above, 
except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
objective function from MR-Egger to get Equation \@ref(eq:lasso) as the needed objective function.

@@

@bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the 
collection of Wald ratio estimates should be homogeneous; thus, any departures from 
homogeneity may signal a violation of one or more assumptions. They reported 
modified weights, in addition to studying weights derived from first order and 
second order approximations to the variance. Their modified weights lead 
to an estimator that is useful for quantifying heterogeneity and detecting outliers.

@@

@foley_mr-clust_2021 recognized that in some MR studies, the Wald ratio estimates for different
SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms 
and may identify the SNPs that are involved in them. They devised a method to detect 
these clusters in the collection of Wald ratio estimates and found SNP instruments that 
point to distinct causal pathways of biological importance [@foley_mr-clust_2021].

MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021]. @tchetgen_tchetgen_genius_2021 focused efforts on relaxing the 
exclusion restriction assumption and building on existing G-estimation 
methods to create an estimator that is robust to violations 
of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting]. They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.
